High-level cognition is supported by information-rich but compressible brain activity patterns

### 2023-12-26 paper/old.tex
submitted version (PNAS R2)

\documentclass[english, 11pt]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage[sort]{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\usepackage{rotating}
\linenumbers

@@

Brain activity patterns are highly flexible and often complex, but also highly
structured. Here we examined how fundamental properties of brain activity
patterns relate to ongoing cognitive processes. To this end, we applied
dimensionality reduction algorithms and pattern classifiers to functional
neuroimaging data collected as participants listened to a story, temporally
scrambled versions of the story, or underwent a resting state scanning session.
These experimental conditions were intended to require different depths of
processing and inspire different levels of cognitive engagement. We considered
two primary aspects of the data. First, we treated the maximum achievable
decoding accuracy across participants as an indicator of the
``informativeness'' of the recorded patterns. Second, we treated the number of
features (components) required to achieve a threshold decoding accuracy as a
proxy for the ``compressibility'' of the neural patterns (where fewer
components indicate greater compression). Overall, we found that the peak
decoding accuracy (achievable without restricting the numbers of features) was
highest in the intact (unscrambled) story listening condition. However, the
number of features required to achieve comparable classification accuracy was
also lowest in the intact story listening condition. Taken together, our work
suggests that our brain networks flexibly reconfigure according to ongoing task
demands, and that the activity patterns associated with higher-order cognition
and high engagement are both more informative and more compressible than the
activity patterns associated with lower-order tasks and lower levels of
engagement.

@@

\bigskip 
\noindent

\textbf{Significance Statement:} How our brains respond to ongoing experiences
depends on what we are doing and thinking about, how engaged we are, and our
past experiences, among other factors. To gain insights into how brain activity
patterns reflect ongoing cognitive processes, we examined two fundamental
aspects of brain activity under different cognitive circumstances:
informativeness and compressibility. Informativeness refers to the extent to
which brain patterns are both temporally specific and consistent across
different people. Compressibility refers to the extent to how robust the
informativeness of brain patterns is to dimensionality reduction. We found that
brain activity patterns evoked by higher-level cognitive tasks are both more
informative \textit{and} more compressible than activity evoked by lower-level
tasks. Our findings suggest that our brains flexibly reconfigure themselves to
optimize different aspects of how they function according to ongoing cognitive
demands.

@@

  \caption{\textbf{Information content and compressibility.} \textbf{A.
  Variance explained for two images.} Consider two example images: a photograph
  and simple drawing. The images have the same resolutions, but their content
  is very different. The photograph is more information rich: when we compare
  the original photograph and drawing (leftmost column), we can see that the
  photograph captures much more detail. We can also apply principal components
  analysis to the images, treating the rows of the images as ``observations.''
  Across columns, we identified the numbers of components required to explain
  100\%, 95\%, 75\%, or 50\% of the cumulative variance in each image (the
  100\% columns denote the original images). The numbers of components are
  indicated in parentheses, and the resulting ``compressed'' images are
  displayed. This analysis reveals that the drawing is more compressible: just
  16 components can explain 95\% of the variance in teh drawing, whereas 50
  components are required to explain 95\% of the variance in the photograph.
  \textbf{B. Representing two images with different numbers of components.}
  Using the same principal component decompositions as in Panel A, we computed
  the cumulative proportion of variance explained with 380 (original images),
  20, 4, or 2 components. This analysis provides another way of characterizing
  the compressibility of each image by showing that the same number of
  components can explain more variance in the drawing than in the photograph.
  \textbf{C. Template data from four synthetic datasets.} We constructed four
  synthetic datasets, each comprising 25 features (columns) observed across 100
  samples (rows) from each of 10 simulated ``participants.'' The datasets were
  constructed to contain different levels of informativeness and
  compressibility (see \textit{Synthetic data}). \textbf{D. Decoding accuracy
  by number of components.} For each synthetic dataset, we trained
  across-participant classifiers to decode timepoint labels.  Each panel displays the
  decoding accuracy as a function of the number of components used to represent
  the data. Error ribbons denote bootstrap-estimated 95\% confidence intervals.}
\label{fig:information-compression} 
\end{figure}

To the extent that brain activity patterns contain rich task-relevant
information, we should be able to use the activity patterns to accurately
differentiate between different aspects of a task~\citep[e.g., using pattern
classifiers;][]{NormEtal06b}. For example, prior work has shown a direct
correspondence between classification accuracy and the information content of a
signal~\citep{Alva02}. To the extent that brain activity patterns are
compressible, we should be able to generate simplified (e.g., lower
dimensional) representations of the data while still preserving the relevant or
important aspects of the original signal. In general, information content and
compressibility are often related but are also dissociable
(Fig.~\ref{fig:information-compression}). If a given signal (e.g., a
representation of brain activity patterns) contains more information about
ongoing cognitive processes, then the peak decoding accuracy should be high. In
the simulations shown in Figure~\ref{fig:information-compression}C we construct
synthetic datasets that have high or low levels of informativeness by varying
temporal autocorrelations in the data (see \textit{Synthetic data}).
If a signal is compressible, then a low-dimensional embedding of the signal
will be similarly informative as the original signal.  In the simulations
shown in Figure~\ref{fig:information-compression}C we construct synthetic
datasets that have high or low levels of compressibility by varying
the covariance structure across features (see \textit{Synthetic data}).  As shown
in Figure~\ref{fig:information-compression}D, highly informative datasets
yield higher decoding accuracies than less informative datasets (i.e., the peaks of the curves in
the top panels of Fig.~\ref{fig:information-compression}D are higher than the
peaks of the curves in the bottom panels). Highly compressible datasets show
steeper slopes when we plot decoding accuracy as a function of the number of
components used to represent the data (i.e., the slopes of the curves in the
left panels of Fig.~\ref{fig:information-compression}D are steeper than the
slopes of the curves in the right panels).  Whereas characterizing
the informativeness and compressibility of synthetic data can be instructive,
we are ultimately interested in understanding how these properties relate to
brain activity patterns recorded under different cognitive circumstances.

Several recent studies suggest that the complexity of brain activity is
task-dependent, whereby simpler tasks with lower cognitive demands are
reflected by simpler and more compressible brain activity patterns, and more
complex tasks with higher cognitive demands are reflected by more complex and
less compressible brain activity patterns~\citep{MackEtal20, OwenEtal21}. These
patterns hold even when the stimulus itself is held
constant~\citep{MackEtal20}. These findings complement other work suggesting
that functional connectivity (correlation) patterns are
task-dependent~\citep{FinnEtal17, OwenEtal20, ColeEtal14}, although
see~\cite{GratEtal18}. Higher-order cognitive processing of a common stimulus
also appears to drive more stereotyped task-related activity and functional
connectivity across individuals~\citep{HassEtal08, LernEtal11, SimoChan20,
SimoEtal16}.

The above studies are consistent with two potential descriptions of how
cognitive processes are reflected in brain activity patterns. One possibility
is that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns decreases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings should carry more information (over and
above the information carried by other features) during complex or high-level
(versus simple or low-level) cognitive tasks. The tradeoffs between these two
aspects of brain activity may also vary across brain regions or networks, for
example according to each region's functional role.

We used a previously collected neuroimaging dataset to estimate the extent to
which each of these two possibilities might hold. The dataset we examined
comprised functional magnetic resonance imaging (fMRI) data collected as
participants listened to an audio recording of a 7-minute story, temporally
scrambled recordings of the story, or underwent a resting state
scan~\citep{SimoEtal16}. Each of these experimental conditions evokes different
depths of cognitive processing~\citep{SimoEtal16,LernEtal11,
HassEtal08,OwenEtal21}. We used across-participant classifiers to decode
listening times in each condition, as a proxy for how ``informative'' the
task-specific activity patterns were~\citep{SimoChan20}. We also used principle
components analysis to generate lower-dimensional representations of the
activity patterns. We then repeated the classification analyses after
preserving different numbers of components and examined how classification
accuracy changed across the different experimental conditions.

@@

We analyzed a dataset collected by \cite{SimoEtal16} that comprised four
experimental conditions. These conditions exposed participants to stimuli that
systematically varied in cognitive engagement. In the \textit{intact}
experimental condition, participants listened to an audio recording of a
7-minute Moth Radio Hour story, \textit{Pie Man}, by Jim O'Grady. In the
\textit{paragraph}-scrambled experimental condition, participants listened to a
temporally scrambled version of the story, where the paragraphs occurred out of
order, but where the same set of paragraphs was presented over the entire
listening interval. All participants in this condition experienced the
scrambled paragraphs in the same order. In the \textit{word}-scrambled
experimental condition, participants listened to a temporally scrambled version
of the story, where the words occurred in a random order. Again, all
participants in this condition experienced the scrambled words in the same
order. Finally, in the \textit{rest} experimental condition, participants lay
in the scanner with no overt stimulus, while keeping their eyes open and
blinking as needed. This public dataset provided a convenient means for testing
our hypothesis that different levels of cognitive processing and engagement
affect how informative and compressible the associated brain patterns are.

@@

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. For each condition, decoding accuracy has been normalized
by subtracting expected ``chance'' decoding performance (estimated as
$\frac{1}{T}$, where $T$ is the number of timepoints in the given condition).
This normalization was intended to enable fairer comparisons across conditions;
a relative decoding accuracy of 0.0 denotes ``chance'' performance. The
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows. All
error ribbons in Panels A and B denote bootstrap-estimated 95\% confidence
intervals. \textbf{C. Estimating ``compressibility'' for each condition.} The
probability density plots display the numbers of components required to reach
at least 5\% (corrected) decoding accuracy, across all cross validation runs.
For the ``rest'' condition, where decoding accuracy never reached 5\% accuracy,
we display the distribution of the numbers of components required to achieve
the peak decoding accuracy in each fold. (This distribution is nearly flat, as
also illustrated in Panel E.) The scale bar denotes a height of 0.01.
\textbf{D. Estimating ``informativeness'' for each condition.} The probability
density plots display the peak (corrected) decoding accuracies achieved in each
condition, across all cross validation runs. The scale bar denotes a height of
0.01. \textbf{E. Informativeness versus compressibility.} Each dot displays the
minimum number of components required to achieve at least 5\% corrected
decoding accuracy ($x$-coordinate; for the rest condition the numbers of
components are estimated using the peak decoding accuracies as in Panel C) and
the peak (corrected) decoding accuracy ($y$-coordinate). The smaller dots
correspond to individual cross validation runs and the larger dots display the
averages across runs. The inset displays a zoomed-in view of the indicated
region in the main panel.}

@@

These results make some intuitive sense. As the contextual information
available to participants increases (i.e., over time in the cognitively rich
intact and paragraph conditions), it makes sense that this might
constrain neural responses to a greater extent. While this pattern may not
necessarily hold for \textit{every} possible story or stimulus, we suspect that
it is generally the case that our knowledge about what is happening in a story
tends to increase as we experience more about it. In turn, this could lead to
greater consistency in different people's interpretations of and neural
responses to the stimulus. Similarly, as participants are left to ``mind
wander,'' or as they experience mental fatigue (i.e., over time in the less
cognitively rich word and rest conditions), we suggest that this might
lead to greater variability in neural responses across people, resulting in
lower decoding accuracy. Again, it is not necessarily the case that every
possible ``unengaging'' stimulus will lead to greater neural variability as
time progresses, but we suspect this phenomenon is likely to hold for a variety
of such stimuli. These findings replicate at least to a limited extent (e.g.,
across both cognitively rich conditions and across both cognitively
impoverished conditions, and for the different groups of participants in each
of those conditions). However, determining whether these patterns generalize to
other stimuli would require additional study (with new stimuli).

@@

If the informativeness and compressibility of brain activity patterns vary over
time, do these properties might also vary across brain networks? We used a
network parcellation identified by \cite{YeoEtal11} to segment the brain into
seven distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Figs.~\ref{fig:networks}A--C):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy tended to
increase in the higher-level experimental conditions and decrease (slightly) in
the lower-level experimental conditions (Fig.~\ref{fig:networks}D, E;
Spearman's rank correlation between decoding accuracy and network order:
intact: $\rho = 0.362, p < 0.001$; paragraph: $\rho = 0.441, p < 0.001$; word:
$\rho = -0.102, p = 0.007$; rest: $\rho = -0.354, p < 0.001$). This suggests
that higher-order networks may carry more content-relevant or stimulus-driven
``information.'' We found no clear trends in the proportions of components
required to achieve 5\% decoding accuracy across networks or conditions
(Fig.~\ref{fig:networks}F). We note that the limbic network we considered here
often overlaps with low (imaging) signal regions, and therefore it may be
difficult to draw strong conclusions about this network's informativeness or
compressibility. We also considered the possibility that the correlations with
network order might be influenced by the numbers of nodes in each network. We
designed a permutation-based procedure to address this possibility, whereby we
repeated the above analyses using shuffled network labels (see \textit{Network
permutation tests}). The correlations between decoding accuracy and network
order were reliably more positive than the shuffled correlations for the intact
($t(1998) = 276.431, p < 0.001$) and paragraph ($t(1998) = 330.334, p < 0.001$)
conditions, and reliably more negative for the word ($t(1998) = -16.386, p <
0.001$) and rest ($t(1998) = -318.631, p < 0.001$) conditions. These results
suggest that the correlations between decoding accuracy and network order were
not driven solely by the numbers of nodes in each network.

\begin{sidewaysfigure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/neurosynth_by_component}

@@

\end{sidewaysfigure}

Whereas the above analyses examinine different networks in isolation, how does
full-brain (i.e., potentially multi-network) activity patterns reflected by
different principal components vary across different experimental conditions?
As shown in Figure~\ref{fig:neurosynth-pca}, we used
Neurosynth~\citep{RubiEtal17} to identify, for each component, the associations
with each of 80 themes (see \textit{Reverse inference}). In general, the first
principal components across all of the experimental conditions tended to weight
most heavily on themes related to cognitive control, memory, language
processing, attention, and perception. Other components appeared to vary more
across conditions.

To gain further insights into which brain functions might be most closely
associated with the top-weighted components from each experimental condition,
we manually grouped each Neurosynth-derived topic into a smaller set of core
cognitive functions. Separately for each component, we computed the average
weightings across all topics that were tagged as being associated with each of
these cognitive functions (Figs.~\ref{fig:neurosynth-summary}A,~\neurosynthFull
A). To help visualize these associations, we used the patterns of associations
for each component to construct graphs whose nodes were experimental conditions
and cognitive functions (Figs.~\ref{fig:neurosynth-summary}B,~\neurosynthFull
B). We also computed correlations between the sets of per-topic weightings from
each of the top-weighted components from each experimental condition
(Fig.~\ref{fig:neurosynth-summary}C, D) and between the brain maps for each
condition's components (Fig.~\neurosynthFull C, D). Taken together, we found
that each component appeared to weight on a fundamental set of cognitive
functions that varied by experimental condition. For example, the top principal
components from every condition weighted similarly (across conditions) on the
full set of Neurosynth topics (Fig.~\ref{fig:neurosynth-pca}A) and cognitive
functions (Figs.~\ref{fig:neurosynth-summary}A, B and~\neurosynthFull~A, B),
suggesting that these components might reflect a set of functions or activity
patterns that are common across all conditions. The second components'
weightings were similar across the intact, paragraph, and rest conditions
(highest-weighted functions: cognitive control, memory, social cognition, and
resting state), but different for the word condition (highest-weighted
functions: sensory perception and cognitive control). The fourth components'
weighting grouped the paragraph and word conditions (highest-weighted
functions: memory, language processing, and cognitive control) and the intact
and rest conditions (highest-weighted functions: emotion, social cognition). We
also used ChatGPT~\citep{ChatGPT} to sort the list of manually tagged cognitive
functions from lowest-level to highest-level (Tab.~\topicTags,
Fig.~\ref{fig:neurosynth-summary}E; also see \textit{Ranking cognitive
processes}). We found that higher-level functions tended to be weighted on more
heavily by top components from the intact and paragraph conditions than
lower-level functions (intact vs. word: $t(198) = 11.059, p < 0.001$; intact
vs. rest: $t(198) = 3.699, p < 0.001$; paragraph vs. word: $t(198) = 13.504, p
< 0.001$; paragraph vs. rest: $t(198) = 4.812, p < 0.001$; also see
\textit{Ranking cognitive processes}). The top components from the word
condition showed the opposite tendency, whereby \textit{lower}-level functions
tended to be weighted on more heavily than higher-level functions (word vs.
rest: $t(198) = -7.315, p < 0.001$). The weighting trends for the intact and
paragraph conditions were not reliably different ($t(198) = -0.479, p =
0.633$). The components from the rest condition showed almost no differences in
the weights associated with high-level versus low-level functions (rest vs. 0:
$t(99) = 1.836, p = 0.081$). These findings suggest that when participants were
engaged more strongly (in the more engaging intact and paragraph conditions),
their dominant neural patterns reflected higher-level cognitive functions. In
contrast, when participants were engaged less strongly (in the less engaging
word and rest conditions), their dominant neural patterns reflected lower-level
cognitive functions. Although they were highly statistically reliable, it is
also important to note that these latter effects are also relatively small
(e.g., the slopes for \textit{all} of the experimental conditions are
numerically close to zero; Fig.~\ref{fig:neurosynth-summary}E). We suggest that
this phenomenon may merit further investigation in future work.

@@

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and several core cognitive function classifications is
displayed in the lower right of the figure. Table~\topics~provides a list of
each topic's top-weighted terms, along with each topic's manually labeled
cognitive classification. A full list of the topics most highly associated with
each component may be found in Figure~\topTerms. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heat maps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heat maps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition components, by
component.} Each sub-panel's network plot summarizes the pattern of
correlations between the $n$\textsuperscript{th} top-weighted principal
components (sub-panel) for each experimental condition (gray-outlined dots).
The line thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heat maps in Panel C). \textbf{E. Change in
weights by condition.} The bar heights reflect the slopes of regression lines,
fit separately to the top five components from each condition, between the
ChatGPT-derived ``rank'' of each cognitive classification and the correlations
between the component and topic maps associated with cognitive processes at the
given rank (see \textit{Ranking cognitive processes}) Error bars denote
bootstrap-estimated 95\% confidence intervals. Also see
Fig.~\neurosynthFull~for additional information.}

@@

Cognitive neuroscientists are still grappling with basic questions about the
fundamental ``rules'' describing how our brains respond, and about how brain
activity patterns and the associated underlying cognitive representations and
computations are linked. We identified two aspects of brain activity patterns,
informativeness and compressibility, that appear to change systematically with
task demands and across brain networks. We speculate that these changes may
reflect ongoing tradeoffs between how robust to signal corruption versus how
expressive about ongoing cognitive states our brains' activity patterns are.
Our work also provides a new framework for evaluating these tradeoffs in other
datasets, or in future studies.

>>>>

\documentclass[english, 11pt]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage[sort]{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\linenumbers

@@

We applied dimensionality reduction algorithms and pattern classifiers to
functional neuroimaging data collected as participants listened to a story,
temporally scrambled versions of the story, or underwent a resting state
scanning session. These experimental conditions were intended to require
different depths of processing and inspire different levels of cognitive
engagement. We considered two primary aspects of the data. First, we treated
the maximum achievable decoding accuracy across participants as an indicator of
the ``informativeness'' of the recorded patterns. Second, we treated the number
of features (components) required to achieve a threshold decoding accuracy as a
proxy for the ``compressibility'' of the neural patterns (where fewer
components indicate greater compression). Overall, we found that the peak
decoding accuracy (achievable without restricting the numbers of features)
was highest in the intact (unscrambled) story listening condition.  However, the
number of features required to achieve comparable classification accuracy
was also lowest in the intact story listening condition.  Taken together, our work
suggests that our brain networks flexibly reconfigure according to ongoing task
demands, and that the activity patterns associated with higher-order cognition
and high engagement are both more informative and more compressible than the
activity patterns associated with lower-order tasks and lower levels of
engagement.

@@

@@

  \caption{\textbf{Information content and compressibility.} \textbf{A.
  Variance explained for two images.} We applied principal components analysis
  to a photograph and drawing, treating the rows of the images as
  ``observations.'' Across columns, we identified the numbers of components
  required to explain 100\%, 95\%, 75\%, or 50\% of the cumulative variance in
  each image (the 100\% columns denote the original images). The numbers of
  components are indicated in parentheses, and the resulting ``compressed''
  images are displayed. \textbf{B. Representing two images with different
  numbers of components.} Using the same principal component decompositions as
  in Panel A, we computed the cumulative proportion of variance explained with
  380 (original images), 20, 4, or 2 components. \textbf{C. Cumulative variance
  explained versus number of components.} For the images displayed in Panels A
  and B, we plot the cumulative proportion of variance explained as a function
  of the number of components used to represent each image. \textbf{D.
  Information rate and compressibility.} Across multiple images, the
  information rate (i.e., the amount of information contained in each image;
  horizontal axis) is high when each individual pixel provides information that
  cannot be inferred from other pixels. High-information rate images tend to be
  high-resolution, and low-information rate images tend to be low-resolution.
  Compressibility is related to the difference between the information required
  to specify the original versus compressed images (vertical axis). Highly
  compressible images often contain predictable structure (redundancies) that
  can be leveraged to represent the images much more efficiently than in their
  original feature spaces.}

\label{fig:information-compression} 
\end{figure}

To the extent that brain activity patterns contain rich task-relevant
information, we should be able to use the activity patterns to accurately
differentiate between different aspects of a task~\citep[e.g., using pattern
classifiers;][]{NormEtal06b}. For example, prior work has shown a direct
correspondence between classification accuracy and the information content of a
signal~\citep{Alva02}. To the extent that brain activity patterns are
compressible, we should be able to generate simplified (e.g., lower
dimensional) representations of the data while still preserving the relevant or
important aspects of the original signal. In general, information content and
compressibility are related but are partially dissociable
(Fig.~\ref{fig:information-compression}). If a given signal (e.g., a
representation of brain activity patterns) contains more information about
ongoing cognitive processes, then the peak decoding accuracy should be high.
And if the signal is compressible, a low-dimensional embedding of the signal
will be similarly informative as the original signal
(Fig.~\ref{fig:information-compression}D).

Several recent studies suggest that the complexity of brain activity is
task-dependent, whereby simpler tasks with lower cognitive demands are
reflected by simpler and more compressible brain activity patterns, and more
complex tasks with higher cognitive demands are reflected by more complex and
less compressible brain activity patterns~\citep{MackEtal20, OwenEtal21}. These
findings complement other work suggesting that functional connectivity
(correlation) patterns are task-dependent~\citep{FinnEtal17, OwenEtal20,
ColeEtal14}, although see~\cite{GratEtal18}. Higher-order cognitive processing
of a common stimulus also appears to drive more stereotyped task-related
activity and functional connectivity across individuals~\citep{HassEtal08,
LernEtal11, SimoChan20, SimoEtal16}.

The above studies are consistent with two potential descriptions of how
cognitive processes are reflected in brain activity patterns. One possibility
is that the information rate of brain activity increases during more complex or
higher-level cognitive processing. If so, then the ability to reliably decode
cognitive states from brain activity patterns should improve with task
complexity or with the level (or ``depth'') of cognitive processing. A second
possibility is that the compressibility of brain activity patterns increases
during more complex or higher-level cognitive processing. If so, then
individual features of brain recordings, or compressed representations of brain
recordings, should carry more information during complex or high-level (versus
simple or low-level) cognitive tasks.

We used a previously collected neuroimaging dataset to estimate the extent to
which each of these two possibilities might hold. The dataset we examined
comprised functional magnetic resonance imaging (fMRI) data collected as
participants listened to an audio recording of a 10-minute story, temporally
scrambled recordings of the story, or underwent a resting state
scan~\citep{SimoEtal16}. Each of these experimental conditions evokes different
depths of cognitive processing~\citep{SimoEtal16,LernEtal11,
HassEtal08,OwenEtal21}. We used across-participant classifiers to decode
listening times in each condition, as a proxy for how ``informative'' the
task-specific activity patterns were~\citep{SimoChan20}. We also used principle
components analysis to generate lower-dimensional representations of the
activity patterns. We then repeated the classification analyses after
preserving different numbers of components and examined how classification
accuracy changed across the different experimental conditions.

@@

We analyzed a dataset collected by \cite{SimoEtal16} that comprised four
experimental conditions. These conditions exposed participants to stimuli that
systematically varied in cognitive engagement. In the \textit{intact}
experimental condition, participants listened to an audio recording of a
10-minute Moth Radio Hour story, \textit{Pie Man}, by Jim O'Grady. In the
\textit{paragraph}-scrambled experimental condition, participants listened to a
temporally scrambled version of the story, where the paragraphs occurred out of
order, but where the same set of paragraphs was presented over the entire
listening interval. All participants in this condition experienced the
scrambled paragraphs in the same order. In the \textit{word}-scrambled
experimental condition, participants listened to a temporally scrambled version
of the story, where the words occurred in a random order. Again, all
participants in this condition experienced the scrambled words in the same
order. Finally, in the \textit{rest} experimental condition, participants lay
in the scanner with no overt stimulus, while keeping their eyes open and
blinking as needed. This public dataset provided a convenient means for testing
our hypothesis that different levels of cognitive processing and engagement
affect how informative and compressible the associated brain patterns are.

@@

\caption{\textbf{Decoding accuracy and compression.} \textbf{A. Decoding
accuracy by number of components.} Ribbons of each color display
cross-validated decoding performance for each condition (intact, paragraph,
word, and rest), as a function of the number of components (features) used to
represent the data. For each condition, decoding accuracy has been normalized
by subtracting expected ``chance'' decoding performance (estimated as
$\frac{1}{T}$, where $T$ is the number of timepoints in the given condition).
This normalization was intended to enable fairer comparisons across conditions;
a relative decoding accuracy of 0.0 denotes ``chance'' performance. The
horizontal black line denotes 5\% decoding accuracy (used as a reference in
Panel B). \textbf{B. Numbers of components required to reach a fixed decoding
accuracy threshold, by condition.} The panel displays a zoomed-in view of the
inset in Panel A. Intersections between each condition's decoding accuracy
curve and the 5\% decoding accuracy reference line are marked by arrows. All
error ribbons in Panels A and B denote bootstrap-estimated 95\% confidence
intervals. \textbf{C. Estimating ``compressibility'' for each condition.} The
density plots display the numbers of components required to reach at least 5\%
(corrected) decoding accuracy, across all cross validation runs. For the
``rest'' condition, where decoding accuracy never reached 5\% accuracy, we
display the distribution of the numbers of components required to achieve the
peak decoding accuracy in each fold. (This distribution is nearly flat, as
also illustrated in Panel E.) \textbf{D. Estimating ``informativeness'' for each
condition.} The density plots display the peak (corrected) decoding accuracies
achieved in each condition, across all cross validation runs. \textbf{E.
Informativeness versus compressibility.} Each dot displays the minimum number
of components required to achieve at least 5\% corrected decoding accuracy
($x$-coordinate; for the rest condition the numbers of components are estimated
using the peak decoding accuracies as in Panel C) and the peak (corrected)
decoding accuracy ($y$-coordinate). The smaller dots correspond to individual
cross validation runs and the larger dots display the averages across runs. The
inset displays a zoomed-in view of the indicated region in the main panel.}

@@

@@

We also wondered how informativeness and compressibility in the different
experimental conditions might vary across brain networks. We used a network
parcellation identified by \cite{YeoEtal11} to segment the brain into seven
distinct networks. The networks can be sorted (roughly) in order from
lower-level to higher-level cortex as follows (Figs.~\ref{fig:networks}A--C):
visual, somatomotor, dorsal attention, ventral attention, limbic,
frontoparietal, and default mode. Next, we computed decoding curves separately
for the activity patterns within each network and identified each network's
inflection point, for each experimental condition. Moving from low-order
networks to higher-order networks, we found that decoding accuracy tended to
increase, particularly in the higher-level experimental conditions
(Fig.~\ref{fig:networks}D, E). This suggests that higher-order networks may
carry more content-relevant or stimulus-driven ``information.'' We found no
clear trends in the proportions of components required to achieve 5\% decoding
accuracy across networks or conditions (Fig.~\ref{fig:networks}F).

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/neurosynth_by_component}

@@

In addition to examining different networks in isolation, we wondered about the
general structure of the full-brain (i.e., potentially multi-network) activity
patterns reflected by different principal components across different
experimental conditions. As shown in Figure~\ref{fig:neurosynth-pca}, we used
Neurosynth~\citep{RubiEtal17} to identify, for each component, the associations
with each of 80 themes (see \textit{Reverse inference}).  In general, the first
principal components across all of the experimental conditions tended to weight
most heavily on themes related to cognitive control, memory,
language processing, attention, and perception.  Other components appeared to
vary more across conditions.

To gain further insights into which brain functions might be most closely
associated with the top-weighted components from each experimental condition,
we manually grouped each Neurosynth-derived topic into a smaller set of core
cognitive functions. Separately for each component, we computed the average
weightings across all topics that were tagged as being associated with each of
these cognitive functions (Figs.~\ref{fig:neurosynth-summary}A,~\neurosynthFull
A). To help visualize these associations, we used the patterns of associations
for each component to construct graphs whose nodes were experimental conditions
and cognitive functions (Figs.~\ref{fig:neurosynth-summary}B,~\neurosynthFull
B). We also computed correlations between the sets of per-topic weightings from
each of the top-weighted components from each experimental condition
(Fig.~\ref{fig:neurosynth-summary}C, D) and between the brain maps for each
condition's components (Fig.~\neurosynthFull C, D). Taken together, we found
that each component appeared to weight on a fundamental set of cognitive
functions that varied by experimental condition. For example, the top principal
components from every condition weighted similarly (across conditions) on the
full set of Neurosynth topics (Fig.~\ref{fig:neurosynth-pca}A) and cognitive
functions (Figs.~\ref{fig:neurosynth-summary}A, B and~\neurosynthFull~A, B),
suggesting that these components might reflect a set of functions or activity
patterns that are common across all conditions. The second components'
weightings were similar across the intact, paragraph, and rest conditions
(highest-weighted functions: cognitive control, memory, social cognition, and
resting state), but different for the word condition (highest-weighted
functions: sensory perception and cognitive control). The fourth components'
weighting grouped the paragraph and word conditions (highest-weighted
functions: memory, language processing, and cognitive control) and the intact
and rest conditions (highest-weighted functions: emotion, social cognition). We
also used ChatGPT~\citep{ChatGPT} to sort the list of manually tagged cognitive
functions from lowest-level to highest-level (Tab.~\topicTags,
Fig.~\ref{fig:neurosynth-summary}E; also see \textit{Ranking cognitive
processes}). We found that higher-level functions tended to be weighted on more
heavily by top components from the intact and paragraph conditions than
lower-level functions. The top components from the word condition showed the
opposite tendency, whereby \textit{lower}-level functions tended to be weighted
on more heavily than higher-level functions. The components from the rest
condition showed almost no differences in the weights associated with
high-level versus low-level functions.

@@

\caption{\textbf{Summary of functions associated with top-weighted components
by condition.} \textbf{A. Top-weighted topics by condition.} Here we display
per-condition (rows, indicated by colored dots) topic correlations, averaged
across topics that pertain to each of several broad cognitive functions
(columns within each sub-panel, indicated by colored dots). Each sub-panel
reflects correlations for the components indicated in the panel titles. A
legend for the condition and several core cognitive function classifications is
displayed in the lower right of the figure. Table~\topics~provides a list of
each topic's top-weighted terms, along with each topic's manually labeled
cognitive classification. A full list of the topics most highly associated with
each component may be found in Figure~\topTerms. \textbf{B. Associations
between per-condition components and cognitive functions.} The network plots
denote positive average correlations between the component images for each
condition (gray-outlined dots on the left sides of each network; colors denote
conditions) and topic-specific brain maps associated with each indicated
cognitive function (black-outlined dots on the right sides of each network;
colors denote cognitive functions). The line thicknesses are proportional to
the correlation values (correlation coefficients are noted in the heat maps in
Panel A). \textbf{C. Correlations between each principal component, by
condition.} The heat maps display the correlations between the brain maps
(Fig.~\componentBrains) for each principal component (sub-panel), across each
pair of conditions (rows and columns of each sub-panel's matrix, indicated by
colored dots). \textbf{D. Associations between per-condition components, by
component.} Each sub-panel's network plot summarizes the pattern of correlations
between the $n$\textsuperscript{th} top-weighted principal components
(sub-panel) for each experimental condition (gray-outlined dots). The line
thicknesses are proportional to the correlation values (correlation
coefficients are noted in the heat maps in Panel C). \textbf{E. Change in
weights by condition.} The bar heights reflect the slopes of regression lines,
fit separately to the top five components from each condition, between the
ChatGPT-derived ``rank'' of each cognitive classification and the correlations
between the component and topic maps associated with cognitive processes at the
given rank (see \textit{Ranking cognitive processes}).  Also see
Fig.~\neurosynthFull~for additional information.}

@@

Cognitive neuroscientists are still grappling with basic questions about the
fundamental ``rules'' describing how our brains respond, and about how brain
activity patterns and the associated underlying cognitive representations and
computations are linked. We identified two aspects of brain activity patterns,
informativeness and compressibility, that appear to change systematically with
task demands and across brain networks. Our work helps to clarify how the
``neural code'' might be structured, and how the code might vary across tasks
and brain areas.
